---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dispositionEffect

<!-- badges: start -->
<!-- badges: end -->

R Package to analyze the Disposition Effect (first formalized by H. Shefrin 
&amp; M. Statman, 1985).  
The goal of dispositionEffect is to ...


## Installation

You can install the released version of dispositionEffect from 
[CRAN](https://CRAN.R-project.org) with:

``` r
# install.packages("dispositionEffect")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
# devtools::install_github("marcozanotti/dispositionEffect")
```

By the moment, since we are still working with a Privae Repo,
you need to install the development version via PAT from the 
appropriate branch.

``` r
install.packages("devtools")
devtools::install_github("marcozanotti/dispositionEffect",
                         ref = "branch_name",
                         auth_token = "cf248fcfa25a81c3caea95896beb28f4e9878ff6")
```


## Example

This is a basic example which shows you how to solve a common problem:

```{r example, eval=FALSE}
library(dispositionEffect)
res <- portfolio_compute(portfolio_transactions = dispositionEffect::investor, 
												market_prices = dispositionEffect::marketprices, 
												method = "all", 
												allow_short = TRUE, 
												time_threshold = "0 mins",
												posneg_portfolios = FALSE, 
												portfolio_statistics = FALSE,
												verbose = c(1, 1), 
												progress = TRUE)
```



