---
title: "The Analysis of Disposition Effect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Analysis of Disposition Effect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```
  
&nbsp;     
  
# The Disposition Effect

In recent years, an irrational phenomenon in financial markets is grabbing
the attention of behavioral economists: the disposition effect. 
Firstly discovered by H. Shefrin and M. Statman (1985), the disposition effect
consists in the realization that investors are more likely to sell an asset 
when it is gaining value compared to when it is losing value. A phenomenon 
which is closely related to sunk costs’ bias, diminishing sensitivity, and 
loss aversion.   
  
From 1985 until now, the disposition effect has been documented in US retail
stock investors as well as in foreign retail investors and even among 
professionals and institutions. By the time, it is a well-established fact that
the disposition effect is a real behavioral anomaly that strongly influences the
final profits (or losses) of investors. Furthermore, being able to correctly 
capture these irrational behaviors timely is even more important in periods of 
high financial volatility as nowadays.   
  
The `dispositionEffect` package allows to quickly evaluate the presence of 
disposition effect’s behaviors of an investor based solely on his transactions 
and the market prices of the traded assets.  
  
&nbsp;     
  
# Package loading

```{r, eval = TRUE}
library(dispositionEffect)
library(dplyr)
library(tidyr)
library(lubridate)
library(skimr)
library(ggplot2)
library(ggridges)
library(gghalves)
```
  
&nbsp;     
  
# Data

The dataset `DEanalysis` is  provided within the package allowing to
reproduce a full analysis on the disposition effect.  

```{r, eval=FALSE}
help("DEanalysis")
```
  
The disposition effect analysis is performed on two fundamental types
of data frames: 

* portfolio transactions, that is all the financial transactions an
investor did during a specific period of time. 
A single transaction is made up of 6 features: the investor id, 
the asset id, the type of the transaction (it can be a buy or a sell), 
the traded quantity, the traded price, and the datetime.  
  
* market prices, that is the prices found on the stock markets for 
each traded asset and each transaction datetimes.  


## Portfolio Transactions

First of all, we need to extract and understand the structure 
of the transaction dataset.   

```{r, eval=TRUE}
trx <- DEanalysis$transactions # transactions
head(trx)
```

```{r, eval=TRUE}
skimr::skim(trx)
```

The portfolio transaction dataset is made up of the six 
fundamental variables described above.  
This real sample dataset contains transactions on 10 investors
on 337 traded assets, from January 2010 until December 2018.  

One important feature is the `type` variable. It states if
a transaction is a "Buy" (B) or a "Sell" (S), and only this
two values are allowed.  

```{r, eval=TRUE}
unique(trx$type)
```

Moreover, as expected, not all the investors are active
on the whole period of analysis.   

```{r, eval=TRUE}
# number of transactions of each investor over years
trx %>% 
  dplyr::mutate(year = lubridate::year(datetime)) %>% 
  dplyr::count(investor, year) %>% 
  dplyr::arrange(year) %>% 
  tidyr::pivot_wider(names_from = year, values_from = n) %>% 
  dplyr::left_join(dplyr::count(trx, investor), by = "investor")
```

Clearly, they have similar number of transactions as a whole, 
but they traded on different years.  


## Market Prices

The market prices dataset needs only to have three variables:
asset, datetime and price.  

```{r, eval=TRUE}
mkt <- DEanalysis$marketprices # market prices
head(mkt)
```

```{r, eval=TRUE}
skimr::skim(mkt)
```
Again, not all the assets prices are available in every year
because we only need the market prices for those assets for
the traded datetime (i.e. the datetime of the transactions
dataset).  

```{r, eval=TRUE}
mkt %>% 
  dplyr::mutate(year = lubridate::year(datetime)) %>% 
  dplyr::count(asset, year) %>% 
  dplyr::arrange(year) %>% 
  tidyr::pivot_wider(names_from = year, values_from = n) %>% 
  head(10)
```
  
&nbsp;     
  
# Disposition Effect: Single Investor Analysis

Now, first proceed to analyse the behaviour of a single investor.  

```{r, eval=TRUE}
# Investor QZ621
trx_QZ621 <- dplyr::filter(trx, investor == "QZ621") # transactions
mkt_QZ621 <- dplyr::filter(mkt, asset %in% unique(trx_QZ621$asset)) # market prices
```


## Gains & Losses Calculation

Based solely on this two data frames it is possible to compute
the so-called realized gains (RG), realized losses (RL), 
paper gains (PG), and paper losses (PL), as defined by 
L. Mazzucchelli et al. (2021).

To sum up the main concepts are the followings:

* Realized Gain / Loss => whenever an investor closes a position 
in its portfolio in gain / loss  

* Paper Gain / Loss => all the open positions at the moment of
the transaction, and all the partially closed positions.

The `portfolio_compute` is the core interface of the package and 
it is used to perform all the gains and losses computations.  

```{r, eval=TRUE}
p_res <- portfolio_compute(portfolio_transactions = trx_QZ621, market_prices = mkt_QZ621)
head(p_res)[, -5]
skimr::skim(p_res)
```

Hence, the result is a data frame containing:  

* the final portfolio of the investor (first 5 variables)  
* gains and losses results of the chosen method (all the other variables),
where RG = Realized Gains, RL = Realized Losses, PG = Paper Gains, and 
PL = Paper Losses.  


## Disposition Effect Computation

Once that gains and losses have been computed, it is finally 
possible to evaluate both the disposition effect of the investor
and of each traded assets, where the disposition effect is defined as:

$$DE = \bigg(\frac{RG}{RG + PG}\bigg) - \bigg(\frac{RL}{RL + PL}\bigg)$$

The DE varies between -1 and 1. Positive DE values show the 
presence of disposition effect irrational behaviours, while
negative values show the presence of opposite disposition
effect behaviours. A value of zero show that no disposition
effect exists.  
  
You almost never want to compute the disposition effect directly
via the `disposition_effect` function, but you will mostly rely
on the quicker and easier `disposition_compute` interface, since
it designed to handle many situations.  

```{r, eval=TRUE}
de <- disposition_compute(gainslosses = p_res)
head(de)
```

As can be seen, `disposition_compute` calculates a value of disposition
effect for each asset. In order to obtain the value of disposition
effect of the investor, one can simply compute an aggregate statistic,
such as the mean or the median, on the assets' values.  
To do this we can simply use once again `disposition_compute` specifying
the desired `aggregate_fun`.  

```{r, eval=TRUE}
disposition_compute(gainslosses = p_res, aggregate_fun = mean, na.rm = TRUE)
```

Moreover, by means of the `disposition_summary` function
it is also easy to summarise the disposition effect behaviour 
of the investor, obtaining common summary statistics.   

```{r, eval=TRUE}
de_stat <- disposition_summary(gainslosses = p_res)
de_stat
```
  
&nbsp;     
  
# Disposition Effect: Testing Different Arguments

Until now, we limited our analysis to the default parameters of 
`portfolio_compute`. However, this function has many different
arguments that can be used both to fine tune the analysis and to
perform more advanced calculations, such as the so-called 
*portfolio driven disposition effect* and the 
*time series disposition effect*.  

Hence, we focus here on the usage of five fundamental 
different arguments.  


## Method

Let's start by the `method` argument. It is probably the most relevant
parameter the user can control since it allows to perform five different 
types of analysis.  

If set to "none", no gains and losses are computed but the investor's
portfolio is updated at every transaction, resulting in the actual 
portfolio of the investors at time T (the end of the period).  

```{r, eval=TRUE}
portfolio_compute(portfolio_transactions = trx_QZ621, market_prices = mkt_QZ621, method = "none") %>% 
  head()
```

If set to one of "count", "total", "value", or "duration", 
gains and losses are computed for the corresponding method.  

```{r, eval=TRUE}
portfolio_compute(portfolio_transactions = trx_QZ621, market_prices = mkt_QZ621, method = "count") %>% 
  head()
portfolio_compute(portfolio_transactions = trx_QZ621, market_prices = mkt_QZ621, method = "total") %>% 
  head()
portfolio_compute(portfolio_transactions = trx_QZ621, market_prices = mkt_QZ621, method = "value") %>% 
  head()
portfolio_compute(portfolio_transactions = trx_QZ621, market_prices = mkt_QZ621, method = "duration") %>% 
  head()
```

In particular:  

* `count` computes gains and losses as simple counts  
* `total` calculates gains and losses as the sum of the quantity  
* `value` measures the expected percentage gains and losses based on prices  
* `duration` calculates the time in hours of held gains and losses.  

Instead, when method is set to "all", then all the four measures are
computed.  

```{r, eval=TRUE}
p_res_all <- portfolio_compute(portfolio_transactions = trx_QZ621, market_prices = mkt_QZ621, method = "all")
skimr::skim(p_res_all)
```

It is important to notice that the disposition effect is 
only meaningful for methods `coount` and `total`. In all 
the other cases the disposition difference is used instead.    

```{r, eval=TRUE}
disposition_compute(gainslosses = p_res_all, aggregate_fun = mean, na.rm = TRUE)
```


## Allow Short

The `allow_short` argument, instead, allows for short selling 
transactions. If set to FALSE, short selling will not be allowed
and no gains or losses will be computed when this happens.  


## Time Threshold

The `time_threshold` argument is a fundamental fine tuning 
parameter. It essentially controls the minimum time distance 
necessary to compute gains and losses.  
By default it is set to "0 mins", implying that gains and losses
are always computed.  
However, this may not be desirable since investor's behaviours 
are not expected to change with very high frequencies.  
Hence, for instance, setting it to "60 mins" states that gains and
losses are calculated only if 60 minutes are passed from the last
transaction.  

```{r, eval=FALSE}
portfolio_compute(
  portfolio_transactions = trx_QZ621, 
  market_prices = mkt_QZ621, 
  time_threshold = "60 mins"
)
```

This parameter is very important also because it allows to 
somewhat filters human operations from machines operations, 
without actually removing them from the analysis.  

Different units may be specified.  


## Exact Market Prices

The argument `exact_market_prices` is set to TRUE by default,
since it is expected that the user provides market prices of
each traded asset for each transaction datetime.  
However, when this is not the case, one may want to set it to
FALSE to allow for non exact market prices. It essentially 
means that the nearest price in time is used.  

```{r, eval=FALSE}
portfolio_compute(
  portfolio_transactions = trx_QZ621, 
  market_prices = mkt_QZ621, 
  exact_market_prices = FALSE
)
```

Note, however, that with `exact_market_prices` set to FALSE, 
unreliable results may be obtained when transactions occur 
with low frequency, since the market prices used as reference
for the calculation may be outdated.  


## Interactivity

The `verbose` and `progress` arguments may be useful for 
interactive use and for very long calculations on large
portfolios of transactions.  

```{r, eval = FALSE}
portfolio_compute(
  portfolio_transactions = trx_QZ621, 
  market_prices = mkt_QZ621, 
  verbose = c(1, 1),
  progress = TRUE
)
```


## Other Arguments

See 
[Portfolio Driven Disposition Effect]() and
[Time Series Disposition Effect](https://marcozanotti.github.io/dispositionEffect/articles/de-timeseries.html)
for a guide on the usage of other, more advanced, arguments.
  
&nbsp;     
  
# Disposition Effect: Multiple Investors Analysis

Although the analysis of disposition effect can be performed simply
on a single investor, the real advantages of this analysis derive from
the capacity to study and understand the behaviours of many different
investors that actively operate on the financial markets.  

Hence, to fully grasp the power of `dispositionEffect` package, we can
proceed to jointly analyse all the 10 investors' transactions that are
available into the `DEanalysis` dataset.  

Furthermore, if you are interested in computing the disposition effect
on large datasets, please see 
[Disposition Effect in Parallel](https://marcozanotti.github.io/dispositionEffect/articles/de-parallel.html)
to understand how the benefits of parallel computing can be exploited
within this framework.  

```{r, eval=TRUE}
# list of transactions separated by investor
trx_list <- trx %>% 
  dplyr::group_by(investor) %>% 
  dplyr::group_split()
```

## Gains & Losses Calculation

This time to calculate gains and losses for each investor's
portfolio we can simply `map` `portfolio_compute` on the list
of transactions, specifying all the other necerrasy arguments
as usual.  

```{r, eval=FALSE}
p_res_full <- purrr::map(trx_list, portfolio_compute, market_prices = mkt)
```

```{r, eval=TRUE, include=FALSE}
load("figures/p_res_full.RData")
```


## Disposition Effect Computation

The same procedure can be used to quickly compute the disposition
effect on each resulting portfolio.  

```{r, eval=TRUE}
de <- purrr::map(p_res_full, disposition_compute) %>% 
  dplyr::bind_rows()
skimr::skim(de)
```
As it is shown, `de` is a data frame containing disposition effect
results (variable `DE_count`) on the 10 investors for all their 337 
traded assets.  
Also the average disposition effects of the investors can be easily obtained,   

```{r, eval=TRUE}
de_mean <- purrr::map(p_res_full, disposition_compute, aggregate_fun = mean, na.rm = TRUE) %>% 
  dplyr::bind_rows() %>% 
  dplyr::arrange(dplyr::desc(DE_count))
de_mean
```

and the disposition effect summary statistics.

```{r, eval=TRUE}
de_stat <- purrr::map(p_res_full, disposition_summary) %>% 
  dplyr::bind_rows()
head(de_stat, 7)
```

It is clearer now that some investors display irrational behaviours
while other don't.

  
## Visual Analysis

The graphical inspection of disposition effect results allows
to easily understand what is going on and to spot possible
interesting behaviours.  

One may want to investigate the overall distribution, or 
the distributions of every statistics obtained.  

```{r, eval=TRUE, fig.width=8, fig.height=5, fig.align='center'}
ggplot(de, aes(x = DE_count)) +
	geom_histogram(aes(y = ..density..), color = "darkblue", fill = "yellow") +
  geom_density(aes(y = ..density..), color = "darkblue", fill = "yellow", alpha = 0.4) +
	scale_x_continuous(limits = c(-1, 1)) +
	theme(
		panel.background = element_rect(fill = "grey92"),
		plot.background = element_rect(fill = "grey85", colour = NA),
		plot.title = element_text(size = 20),
		legend.position = "none"
	) +
	labs(title = "Disposition Effect Distribution", x = "Disposition Effect", y = "Frequency")
```

```{r, eval=TRUE, include=FALSE}
de_stat <- de_stat %>%
  dplyr::filter(stat != "StDev") %>% 
  dplyr::mutate(stat = factor(stat, levels = c("Min", "Q1", "Median", "Mean", "Q3", "Max")))
```

```{r, eval=TRUE, fig.width=8, fig.height=5, fig.align='center'}
ggplot(de_stat, aes(x = DE_count, y = stat, fill = stat)) +
	geom_density_ridges() +
  scale_fill_viridis_d() +
	scale_x_continuous(limits = c(-1, 1)) +
	theme(
		panel.background = element_rect(fill = "grey92"),
		plot.background = element_rect(fill = "grey85", colour = NA),
		plot.title = element_text(size = 20),
		legend.position = "none"
	) +
	labs(
		title = "Disposition Effect Statistics' Distributions",
		x = "Disposition Effect", y = ""
	)
```

Or deeper, one can also analyse investors' behaviours on some
specific assets to understand whether there exists on the market
assets that are more subject to irrationality.    

```{r, eval=TRUE, fig.width=8, fig.height=5, fig.align='center'}
top5_assets <- trx %>% 
  dplyr::count(asset) %>% 
  dplyr::arrange(dplyr::desc(n)) %>% 
  dplyr::slice(1:6) %>% 
  dplyr::pull(asset)

dplyr::filter(de, asset %in% top5_assets) %>% 
  ggplot(aes(x = asset, y = DE_count, fill = asset)) +
	geom_half_boxplot(center = TRUE, width = 0.8, nudge = 0.02) +
	geom_half_violin(side = "r", nudge = 0.02, alpha = 0.8) +
	geom_jitter(color = "grey40") +
  scale_fill_viridis_d() +
	theme(
		panel.background = element_rect(fill = "grey92"),
		plot.background = element_rect(fill = "grey85", colour = NA),
		plot.title = element_text(size = 20),
		legend.position = "none"
	) +
	labs(title = "Volatility & Disposition Effect", x = "",	y = "Disposition Effect")
```

&nbsp;  

---------------------------------------------------------------------------

For more tutorials on disposition effect visit 
[dispositionEffect](https://marcozanotti.github.io/dispositionEffect).
