---
title: "Time Series Disposition Effect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time Series Disposition Effect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```


# Installation

By the moment, you can only install the development version from
[GitHub](https://github.com/) with:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("marcozanotti/dispositionEffect")
```

You will be able to install the released version of `dispositionEffect`
from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("dispositionEffect")
```
  
  
# Package loading

The following simple example shows how to perform disposition effect 
calculations on real financial market data.  

```{r, eval = TRUE}
library(dispositionEffect)
```
  
  
# Data exploration

Portfolio of transactions of a real investor.

```{r, eval = TRUE}
head(investor)
```

Dataset of market prices of the traded assets.

```{r, eval = TRUE}
head(marketprices)
```


# Gains and Losses

Compute realized gains, realized losses, paper gains and paper losses.

```{r, eval = TRUE}
portfolio_results <- portfolio_compute(
	portfolio_transactions = investor, 
	market_prices = marketprices
)
```


# Disposition Effect

Compute the disposition effect with different methods.

```{r, eval = TRUE}
de <- disposition_compute(portfolio_results)
de
```

Summarise the behaviour of the investor.

```{r, eval = TRUE}
disposition_summary(portfolio_results)
```

Plot the results to spot the presence of the disposition effect.

```{r, eval = TRUE}
library(ggplot2)
ggplot(de, aes(x = asset, y = DE_count, fill = asset)) +
	geom_col() +
	scale_fill_viridis_d() +
	labs(
		title = "Disposition Effect results for the traded assets.",
		subtitle = "Method Count",
		x = "", y = ""
	)
```
