---
title: "Time Series Disposition Effect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time Series Disposition Effect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```
  
&nbsp;  

# The importance of time

The disposition effect consists in the realization that investors
are more likely to sell an asset when it is gaining value compared 
to when it is losing value. A phenomenon which is closely related 
to sunk costs’ bias, diminishing sensitivity, and loss aversion.  

This irrational phenomenon has strong implications on financial 
markets. In particular, it is a violation of the well-known 
"efficient market hypothesis" that bases its foundations on the
theory of rational agents.  
  
Hence, timely capturing and understanding irrational behaviours 
on the financial markets is of primary interest both of 
researchers and investors. For this reason, the `dispositionEffect`
package allows to quickly collect both aggregate and time series
results of the disposition effect, allowing to deeply study
the evolution in time of irrationalities.  
  
&nbsp;  
  
# Package loading

To load the  package simply use the usual `library` function.  

```{r, eval = TRUE}
library(dispositionEffect)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
```
  
&nbsp;  
  
# Data

The disposition effect analysis is performed on two fundamental types
of data frames:

* portfolio transactions, that is all the financial transactions an
investor did during a specific period of time. 
A single transaction is made up of 6 features: the investor id, 
the asset id, the type of the transaction (it can be a buy or a sell), 
the traded quantity, the traded price, and the datetime.

* market prices, that is the prices found on the stock markets for 
each traded asset and each transaction datetimes.  
  
&nbsp;     

# Time series computations

The `portfolio_compute` function is the core interface of the package 
and it is used to perform all the gains and losses computations.  
In particular, the argument `time_series_DE` is used to enable time 
series disposition effect computations. 

```{r, eval = TRUE}
portfolio_results_ts <- portfolio_compute(
	portfolio_transactions = investor, 
	market_prices = marketprices,
	time_series_DE = TRUE
)
```

Setting it to `TRUE` makes the function return two different results:  

* the usual aggregate portfolio with gains and losses results  

```{r, eval=TRUE}
portfolio <- portfolio_results_ts$portfolio
dplyr::select(portfolio, -datetime)
```

* the new time series disposition effect based on the chosen `method`

```{r, eval=TRUE}
timeseries <- portfolio_results_ts$timeseries
head(timeseries)
```

For every transaction datetime two different disposition effect
are computed:  

* DEts, disposition effect on gains and losses at time t  
* DETs, disposition effect on gains and losses at time t taking
into account all the previous results  

Note that, by the moment, the time series computations of disposition
effect are allowed for `"count"` and `"value"` methods only (with 
the latter the disposition difference is implemented instead of 
disposition effect).  
  
&nbsp;     

# Investor's time series disposition effect

The `disposition_summary_ts` function can be used to summarise
the evolution of disposition effect over time.  

```{r, eval=TRUE}
disposition_summary_ts(timeseries)
```

A visual time series analysis can be performed as usual
with [`ggplot2`](https://ggplot2.tidyverse.org/).  

```{r, eval=TRUE, fig.width=8, fig.height=5, fig.align='center'}
timeseries %>% 
  tidyr::pivot_longer(cols = dplyr::starts_with("DE")) %>% 
  ggplot2::ggplot(ggplot2::aes(x = datetime, y = value, col = name)) +
  ggplot2::geom_line(size = 1.5) +
  ggplot2::scale_colour_viridis_d(alpha = 1) +
	ggplot2::labs(
		title = "Time Series Disposition Effect results",
		subtitle = "Method Count",
		x = "", y = ""
	) +
  ggplot2::theme(legend.position = "bottom")
```
  
&nbsp;   

# Assets' time series disposition effect

The time series analysis of disposition effect can be
greatly improved also by computing the disposition effect
over time on specif assets traded by the investor.  
Indeed, the argument `assets_time_series_DE` allows to specify
a character vector of assets' id (that must be traded by the
investor) on which to compute the disposition effect over time.  

```{r, eval=TRUE}
portfolio_results_ts_assets <- portfolio_compute(
	portfolio_transactions = investor, 
	market_prices = marketprices,
	time_series_DE = TRUE,
	assets_time_series_DE = c("ACO", "LSUG")
)
```

```{r, eval=TRUE}
timeseries_assets <- portfolio_results_ts_assets$timeseries
head(timeseries_assets)[, 2:6]
head(timeseries_assets)[, c(2:4, 7:8)]
```
 
```{r, eval=TRUE}
disposition_summary_ts(timeseries_assets)[, 2:6]
disposition_summary_ts(timeseries_assets)[, c(2:4, 7:8)]
``` 
 
```{r, eval=TRUE, fig.width=8, fig.height=5, fig.align='center'}
timeseries_assets %>% 
  tidyr::pivot_longer(cols = dplyr::contains("DE")) %>% 
  ggplot2::ggplot(ggplot2::aes(x = datetime, y = value, col = name)) +
  ggplot2::geom_line(size = 1.5) +
  ggplot2::scale_colour_viridis_d(alpha = 1) +
  ggplot2::facet_wrap(~ name, ncol = 2) +
	ggplot2::labs(
		title = "Assets Time Series Disposition Effect results",
		subtitle = "Method Count",
		x = "", y = ""
	) +
  ggplot2::theme(legend.position = "bottom")
``` 
 
This way it is possible to better understand the behaviour
of an investor on his traded assets. It may be possible, 
in practice that, disposition effect behaviours are only
present on some specific assets, and understanding what assets
are more subject to irrationality can be critical.  
  
&nbsp;   

# Compare investors' behaviours

In order to better understand the relevance of the time 
series analysis of disposition effect, we can test the
results on the `DEanalysis` real sample dataset.  

See ["Disposition Effect in Parallel"](https://marcozanotti.github.io/dispositionEffect/articles/de-parallel.html) 
to speed up computations with parallel computing in R.  

```{r, eval=FALSE}
trx <- DEanalysis$transactions
mkt <- DEanalysis$marketprices
investor_id <- unique(trx$investor)

res_list <- vector(mode = "list", length = length(investor_id))
for (i in seq_along(investor_id)) {
  tmp_trx <- trx %>%
    dplyr::filter(investor == investor_id[i])
  tmp_res <- tryCatch(
    dispositionEffect::portfolio_compute(
      portfolio_transactions = tmp_trx,
      market_prices = mkt,
      time_series_DE = TRUE
    ),
    error = function(e) "Error"
  )
  res_list[[i]] <- tmp_res # save results
  rm(tmp_trx, tmp_res)
}

# extract time series results for each investor
timeseries_10_investors <- res_list %>% 
  purrr::map("timeseries")
```

```{r, eval=TRUE, include=FALSE}
load("figures/ts_res.RData")
```

```{r, eval=TRUE}
purrr::map(timeseries_10_investors, disposition_summary_ts) %>% 
  dplyr::bind_rows() %>% 
  dplyr::filter(stat == "Mean") %>% 
  dplyr::arrange(desc(DETs_count))
```

```{r, eval=TRUE, fig.width=8, fig.height=5, fig.align='center'}
timeseries_10_investors %>% 
  dplyr::bind_rows() %>%
  tidyr::pivot_longer(cols = dplyr::contains("DE")) %>% 
  ggplot2::ggplot(ggplot2::aes(x = datetime, y = value, col = investor)) +
  ggplot2::geom_line(size = 0.75) +
  ggplot2::scale_colour_viridis_d(alpha = 0.9) +
  ggplot2::facet_wrap(~ name, nrow = 2, ncol = 1) +
	ggplot2::labs(
		title = "10 Investors Time Series Disposition Effect results",
		subtitle = "Method Count",
		x = "", y = ""
	) +
  ggplot2::theme(legend.position = "bottom")
```

&nbsp;  

---------------------------------------------------------------------------

For more tutorials on disposition effect visit 
[dispositionEffect](https://marcozanotti.github.io/dispositionEffect).
