---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```
  
&nbsp;  

# The Disposition Effect

In recent years, an irrational phenomenon in financial markets is grabbing
the attention of behavioral economists: the disposition effect. 
Firstly discovered by H. Shefrin and M. Statman (1985), the disposition effect
consists in the realization that investors are more likely to sell an asset 
when it is gaining value compared to when it is losing value. A phenomenon 
which is closely related to sunk costs’ bias, diminishing sensitivity, and 
loss aversion.   
  
From 1985 until now, the disposition effect has been documented in US retail
stock investors as well as in foreign retail investors and even among 
professionals and institutions. By the time, it is a well-established fact that
the disposition effect is a real behavioral anomaly that strongly influences the
final profits (or losses) of investors. Furthermore, being able to correctly 
capture these irrational behaviors timely is even more important in periods of 
high financial volatility as nowadays.   
  
The `dispositionEffect` package allows to quickly evaluate the presence of 
disposition effect’s behaviors of an investor based solely on his transactions 
and the market prices of the traded assets.  
  
&nbsp;     
  
# Installation & Loading

By the moment, you can only install the development version from
[GitHub](https://github.com/) with:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("marcozanotti/dispositionEffect")
```

You will be able to install the released version of `dispositionEffect`
from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("dispositionEffect")
```

To load the  package simply use the usual `library` function.  

```{r, eval = TRUE}
library(dispositionEffect)
```
  
&nbsp;     
  
# Data

The disposition effect analysis is performed on two fundamental types
of data frames:

* portfolio transactions, that is all the financial transactions an
investor did during a specific period of time. 
A single transaction is made up of 6 features: the investor id, 
the asset id, the type of the transaction (it can be a buy or a sell), 
the traded quantity, the traded price, and the datetime.

```{r, eval = TRUE}
head(investor)
```

* market prices, that is the prices found on the stock markets for 
each traded asset and each transaction datetimes.

```{r, eval = TRUE}
head(marketprices)
```
  
&nbsp;     
  
# Gains and Losses

Based solely on this two data frames it is possible to compute
the so-called realized gains (RG), realized losses (RL), 
paper gains (PG), and paper losses (PL), as defined by 
L. Mazzucchelli et al. (2021).

To sum up the main concepts are the followings:

* Realized Gain / Loss => whenever an investor closes a position 
in its portfolio in gain / loss  

* Paper Gain / Loss => all the open positions at the moment of
the transaction, and all the partially closed positions.

The `portfolio_compute` is the core interface of the package and 
it is used to perform all the gains and losses computations.

```{r, eval = TRUE}
portfolio_results <- portfolio_compute(
	portfolio_transactions = investor, 
	market_prices = marketprices,
	method = "count"
)
dplyr::select(portfolio_results, -datetime)
```

The result is a new data frame containing RG, PG, RL, and PL for 
each traded asset and the updated investor's portfolio.  
  
&nbsp;     

# Disposition Effect

Once that gains and losses have been computed, it is finally 
possible to evaluate both the disposition effect of the investor
and each traded assets, where the disposition effect is defined as:

$$DE = \bigg(\frac{RG}{RG - PG}\bigg) - \bigg(\frac{RL}{RL - PL}\bigg)$$

The DE varies between -1 and 1. Positive DE values show the 
presence of disposition effect irrational behaviours, while
negative values show the presence of opposite disposition
effect behaviours. A value of zero show that no disposition
effect exists.

The function `disposition_effect` allow us to compute it.

```{r, eval = TRUE}
# assets' disposition effects
disposition_effect(
	realized_gains = portfolio_results$RG_count, 
	paper_gains = portfolio_results$PG_count, 
  realized_losses = portfolio_results$RL_count, 
	paper_losses = portfolio_results$PL_count
)
```

```{r, eval = TRUE}
# investor's disposition effect
disposition_effect(
	realized_gains = portfolio_results$RG_count, 
	paper_gains = portfolio_results$PG_count, 
  realized_losses = portfolio_results$RL_count, 
	paper_losses = portfolio_results$PL_count
) %>% 
	mean(na.rm = TRUE)
```
  
However, one may prefer to use the `disposition_compute` function
directly on the results obtained from `portfolio_compute` for 
quick and clean computations.

```{r, eval = TRUE}
# assets' disposition effects
asset_de <- disposition_compute(gainslosses = portfolio_results)
asset_de
```

```{r}
# investor's disposition effect
investor_de <- disposition_compute(gainslosses = portfolio_results, aggregate_fun = mean, na.rm = TRUE)
investor_de
```

Moreover, by means of the `disposition_summary` function
it is also easy to summarise the disposition effect behaviour 
of the investor.

```{r, eval = TRUE}
# investor's disposition effect summary statistics
disposition_summary(gainslosses = portfolio_results)
```

&nbsp;     
  
# Plotting

Finally, by means of [`ggplot2`](https://ggplot2.tidyverse.org/), 
stunning charts can be produced to easily spot the presence of 
the disposition effect.  

```{r, eval = TRUE, fig.width=8, fig.height=5, fig.align='center'}
library(ggplot2)

asset_de %>% 
	ggplot2::ggplot(ggplot2::aes(x = asset, y = DE_count, fill = asset)) +
	ggplot2::geom_col() +
	ggplot2::scale_fill_viridis_d() +
	ggplot2::labs(
		title = "Disposition Effect results of the traded assets",
		subtitle = "Method Count",
		x = "", y = ""
	)
```

---------------------------------------------------------------------------

For more detailed tutorials on disposition effect visit 
[dispositionEffect](https://marcozanotti.github.io/dispositionEffect).
